---
title: "Naples Reef (PELEGRI)"
author: "Patrick Pelegri-O'Day"
date: "10/24/2021"
output: html_document
---

### Overview
[Insert text]

**Data:** Reed, D. 2020. SBC LTER: Reef: Long-term experiment: Kelp removal: Urchin size frequency distribution ver 20. Environmental Data Initiative. https://doi.org/10.6073/pasta/5a1e9ef03aa47bd2225c0bb98a02a63b. 

**Data description:** Data contain observations of red and purple urchins observed at 5 locations along the Santa Barbara coast. Data have been updated from their raw version for this assignment. See complete description and metadata (here)[https://sbclter.msi.ucsb.edu/data/catalog/package/?package=knb-lter-sbc.26]

### Setup

Attach packages: `tidyverse`, `here`, `janitor`, `lubridate`, `kableExtra`
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

Read in the data
```{r}
urchins <- 
  read_csv(show_col_types = FALSE, here("data", "urchins.csv"))
```

### Data wrangling

Only include Naples Reef observations for purple urchins observed in 2010
```{r}
purple_urchins_naples <- urchins %>% 
  clean_names() %>% # apply `clean_names()` to convert column names to lower_snake_case
  mutate(date = mdy(date)) %>%  # replace original date column with new date column in YYYY-MM-DD format in which column class is "Date"
  mutate(year = year(date)) %>% # create new column called year that contains the year, derived from the date column
  filter(year == "2010") %>%  # only retain observations where year is 2010
  filter(common_name == "Purple Urchin") %>%  # only retain observations where species common name is Purple Urchin
  filter(site == "NAPL") # only retain observations where site is Naples Reef, abbreviated as NAPL
```

### Exploring urchin data visually

Explore data with faceted Q-Q plots
```{r}
ggplot(data = purple_urchins_naples, aes(sample = size, color = treatment)) +
  geom_qq(geom = "point") +
  geom_qq_line() + 
  facet_wrap(~treatment)
```

Explore data with faceted histogram plots
```{r}
ggplot(data = purple_urchins_naples, aes(x = size, fill = treatment)) +
  geom_histogram() +
  facet_wrap(~treatment)
```

"Annual" means that the urchin was in a treatment group, and "control" means the urchin was in a control group
```{r}
naples_urchins_summary <- purple_urchins_naples %>% # start with purple_urchins_naples
   group_by(treatment) %>% # group by whether the urchin was in a treatment site or not
   summarize(mean = mean(size, na.rm = TRUE),
             sd = sd(size, na.rm = TRUE),
             median = median(size, na.rm = TRUE),
             sample_size = n()
   )

kable(naples_urchins_summary,
      col = c("Treatment", "Mean Size (cm)", "Standard deviation", "Median Size (cm)", "Sample Size"),
      caption = "Caption") %>% 
  kable_styling(full_width = FALSE)
```

