---
title: "Naples Reef (PELEGRI)"
author: "Patrick Pelegri-O'Day"
date: "10/24/2021"
output: html_document
---

### Overview
This document wrangles then explores data on purple urchins in Naples Reef. The urchins are divided into treatment and control groups. Summary statistics related to central tendency and spread of urchin size are presented.

**Data:** Reed, D. 2020. SBC LTER: Reef: Long-term experiment: Kelp removal: Urchin size frequency distribution ver 20. Environmental Data Initiative. https://doi.org/10.6073/pasta/5a1e9ef03aa47bd2225c0bb98a02a63b. 

**Data description:** Data contain observations of red and purple urchins observed at 5 locations along the Santa Barbara coast. Data have been updated from their raw version for this assignment. See complete description and metadata [here](https://sbclter.msi.ucsb.edu/data/catalog/package/?package=knb-lter-sbc.26)

### Setup

Attach packages
```{r setup, include=TRUE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

library(tidyverse)
library(here)
library(janitor)
library(lubridate)
library(kableExtra)
```

Read in the data
```{r}
urchins <- 
  read_csv(show_col_types = FALSE, here("data", "urchins.csv"))
```

### Data wrangling

Only include Naples Reef observations for purple urchins observed in 2010
```{r}
purple_urchins_naples <- urchins %>% 
  clean_names() %>% # apply `clean_names()` to convert column names to lower_snake_case
  mutate(date = mdy(date)) %>%  # replace original date column with new date column in YYYY-MM-DD format in which column class is "Date"
  mutate(year = year(date)) %>% # create new column called year that contains the year, derived from the date column
  filter(year == "2010") %>%  # only retain observations where year is 2010
  filter(common_name == "Purple Urchin") %>%  # only retain observations where species common name is Purple Urchin
  filter(site == "NAPL") # only retain observations where site is Naples Reef, abbreviated as NAPL
```

### Exploring urchin data visually

Explore data with faceted Q-Q plots
```{r}
ggplot(data = purple_urchins_naples, aes(sample = size, color = treatment)) + # define dataset and variables for plot
  geom_qq(geom = "point") + # create Q-Q plot
  geom_qq_line() + # run line of best fit through Q-Q plot
  facet_wrap(~treatment) # facet the graph according to whether the data are from treatment or control group
```

Explore data with faceted histogram plots
```{r}
ggplot(data = purple_urchins_naples, aes(x = size, fill = treatment)) + # define dataset and variables for plot
  geom_histogram() + # make histogram graph
  facet_wrap(~treatment) # facet the graph according to whether the data are from treatment or control group
```

### Create summary table for purple urchin treatment and control groups

Create data frame that stores descriptive statistic values
```{r}
# Note: "Annual" means that the urchin was in a treatment group, and "control" means the urchin was in a control group

naples_urchins_summary <- purple_urchins_naples %>% # start with purple_urchins_naples
   group_by(treatment) %>% # group by whether the urchin was in a treatment site or not
   summarize(mean = round(mean(size, na.rm = TRUE), 1), # find mean size, round to 1 decimal
             sd = round(sd(size, na.rm = TRUE), 1), # find standard deviation based on size, round to 1 decimal
             median = median(size, na.rm = TRUE), # find median size
             sample_size = n() # find sample size
   )
```

Create summary table that visualizes those descriptive statistics
```{r}
kable(naples_urchins_summary,
      col = c("Treatment", "Mean Size (cm)", "Standard deviation", "Median Size (cm)", "Sample Size"),
      caption = "Table 1: Summary statistics for Purple Urchins observed in Naples Reef, comparing treatment (ANNUAL) and control (CONTROL) groups") %>% 
  kable_styling(full_width = FALSE)
```

